\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage[margin=1in]{geometry}
% \usepackage{microtype}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\CAlg}{\mathrm{CAlg}}
\newcommand{\CSpec}{\mathrm{CSpec}}
\newcommand{\CC}{\mathrm{CC}}
\newcommand{\Int}{\mathrm{Int}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\Hol}{\mathrm{Hol}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lv}[1]{\textup{[\texttt{#1}]}}

\hypersetup{
 colorlinks=true,
 linkcolor=blue!60!black,
 citecolor=blue!60!black,
 urlcolor=blue!60!black
}

\title{\textsc{Causal-Algebraic Geometry}:\\[6pt]
A Grothendieck-Type Framework for Partial Orders\\[4pt]
{\large with Applications to Manifold Detection and Arithmetic}}

\author{Thomas DiFiore}

\date{March 2026}

\begin{document}
\maketitle

\begin{abstract}
We develop an algebraic-geometric framework for finite partial orders.
Starting from a \emph{causal algebra}---an associative algebra equipped with a partial order on its idempotents satisfying a causality axiom---we construct a spectrum $\CSpec$, equip it with a structure presheaf of corner algebras (satisfying locality but not gluing; Proposition~\ref{prop:gluing}), build a simplicial cohomology theory on the order complex, and introduce the \emph{Noetherian ratio} $\gamma$, a combinatorial invariant measuring geometric regularity.
We prove a \emph{separation theorem}: two explicitly constructed 5-element posets agreeing on eight classical invariants (width, height, Hasse edges, number of antichains, maximal chains, comparable pairs, order ideals, and join-irreducibles) have different Noetherian ratios ($\gamma = 13/5$ vs $14/5$), establishing that $\gamma$ carries structural information beyond all tested cardinal invariants of finite posets.
We prove that $\CSpec$ is the \emph{terminal object} in the category of causal topologies, reversing the usual Grothendieck logic: the ring structure of the corner algebras forces the topology (the collection of opens).
We establish that $\gamma(\text{chain}) < \gamma(\text{grid}) < \gamma(\text{antichain})$ for all $m \geq 2$ (proved). A universal \emph{width-one bound} $\gamma \leq 2$ holds for all total orders ($\gamma < 2$ for $n \geq 2$). A \emph{width scaling law}:
\[
2^w \;\leq\; |\CC(C)| \;\leq\; (\lceil n/w \rceil^2 + 1)^w,
\]
giving $\gamma = O(n^{2(w-1)})$ for fixed width~$w$ (matching computational data for $w = 1, 2, 3$).
An initial conjecture $\gamma \leq 2^w$ was disproved at $n = 6$ during the formalization process; the correct polynomial bound establishes a dichotomy between bounded-width posets (polynomial $\gamma$) and linear-width posets (exponential $\gamma$), with manifold-like causal sets falling in an intermediate regime.
A non-abelian gauge connection on the causal spectrum yields exact rational Wilson loop values on discretized cylinders; we prove that $\operatorname{tr}(\Hol) = n_{\mathrm{time}} - 1$ independent of spatial circumference.
Applied to the truncated divisibility lattice $(\Z^+ \cap [1,N], \mid)$, prime-generated upsets recover the primes $p \leq N$ as $\CSpec$ points, with $\CSpec$ strictly richer than the classical spectrum (8~points vs 2~prime ideals for $\mathrm{div}(12)$). The number-theoretic M\"obius function is recovered exactly.
A \emph{sheaf-$\gamma$ bridge theorem} proves that the causally convex subsets are exactly the ring homomorphism opens of the structure sheaf, so that $\gamma$ equals the ring homomorphism density.
The ratio $\kappa = 1 - 1/\gamma$ measures the fraction of ring homomorphism opens that are not intervals.
The map from posets to convex structures is many-to-one, extracting geometric content from combinatorial data.
The algebraic constructions are formally verified (source available at~\cite{leanrepo}); Dilworth's theorem is proved in full.
\end{abstract}

\medskip

\noindent\textit{2020 Mathematics Subject Classification.}
06A11, 14A15, 18F20, 55N30, 05E45, 11A25, 60C05.

\smallskip

\noindent\textit{Key words and phrases.}
Causal algebra, non-commutative spectrum, Grothendieck sheaf, Noetherian ratio, holonomy, M\"obius function, manifold detection, random partial orders, poset limits.

\bigskip


%% ================================================================
\section{Introduction}
%% ================================================================

The programme of algebraic geometry, in its most general form, associates to an algebraic object a geometric space equipped with a structure sheaf and extracts invariants via cohomology.
Grothendieck's theory of schemes~\cite{grothendieck1960} accomplished this for commutative rings; Connes' non-commutative geometry~\cite{connes1994} extended the programme to operator algebras; Voevodsky's motivic homotopy theory~\cite{voevodsky2003} bridged algebraic geometry and stable homotopy theory.
Each instance opened new territory by providing a geometric language for algebraic structures that previously lacked one.

This paper extends the programme to partial orders.
A \emph{causal set}---a locally finite partially ordered set---is the mathematical model of discrete spacetime proposed by Bombelli, Lee, Meyer, and Sorkin~\cite{bombelli1987}.
The combinatorial theory of random partial orders in this setting is surveyed by Brightwell and Luczak~\cite{brightwell2015}, who establish that the height of a Poisson process in a box space of dimension~$d$ scales as $n^{1/d}$~\cite{bollobas1991}, that every convergent sequence of classical sequential growth models has a semiorder as its limit~\cite{brightwell2010}, and that the uniform measure on 2-dimensional orders is contiguous with a Poisson process in $\mathbb{M}^2$~\cite{brightwell2008}.
The Benincasa--Dowker action~\cite{benincasa2010} provides a discrete Einstein--Hilbert functional via interval counting and the M\"obius function.
What has been missing is algebraic-geometric infrastructure---a spectrum, a structure presheaf, morphisms---adapted to the non-commutative setting of causal structures.

\textbf{Relation to existing work on convex subsets of posets.}
Interval convexity in posets is a classical notion: a subset $S$
is convex if whenever $a \leq c \leq b$ with $a, b \in S$, then
$c \in S$. The lattice of convex sublattices was studied by
Lihov\'a~\cite{lihova1999} and others; convex geometries
(closure systems satisfying the anti-exchange axiom) were
developed by Edelman and Jamison~\cite{edelmanjamison1985}.
Counting convex subsets in products of chains was initiated by
Barnette, Nichols, and Richmond~\cite{barnette2019}, who gave
explicit formulas for small fixed column width.
Prime ideals of incidence algebras are treated in
Spiegel and O'Donnell~\cite{spiegel1997}.
Our contributions begin where this literature stops:
(a)~the observation that the convex subset count
grows exponentially and detects dimension (the dimension law,
proved for all $d \geq 2$);
(b)~the algebraic characterization of convex subsets as the
ring homomorphism opens of a structure sheaf (the bridge theorem);
(c)~the construction of $\CSpec$ with its universal property;
(d)~the Noetherian ratio $\gamma$ as a manifold-likeness
invariant;
and (e)~the holonomy construction distinguishing posets with
identical classical invariants.

The central obstruction is immediate.
The natural algebra of a partial order is the incidence algebra of Rota~\cite{rota1964}, which is non-commutative: the causality axiom makes it upper-triangular.
Standard non-commutative primality degenerates when applied to such algebras: the zero ideal of the causal subalgebra fails to be NC-prime because causally unrelated elements provide automatic witnesses (${\bf e}_\alpha M {\bf e}_\beta = 0$ for all causal $M$ when $\alpha \| \beta$).
We resolve this by introducing \emph{causal primality} (\S\ref{sec:cspec}), a weakening that restricts the primality test to causally connected triples, and prove that the resulting spectrum $\CSpec$ admits an injective map from the original poset into its points (\S\ref{sec:recovery}).
We then prove that causal primality satisfies a \emph{universal property}: $\CSpec$ is the terminal object in the category of causal topologies (\S\ref{sec:uniqueness}).
In Grothendieck's theory, one defines the spectrum and then constructs the sheaf; here, the logic reverses\textemdash the ring structure of the corner algebras forces the topology (the collection of opens).

The principal application is to the \emph{Hauptvermutung} of causal set theory: the question, open since the founding of the programme in 1987, of whether the causal order alone determines the manifold~\cite{sorkin2003, malament1977, brightwell2015}.
Malament~\cite{malament1977} proved topologically that continuous timelike curves determine spacetime topology up to a conformal factor; Hawking, King, and McCarthy~\cite{hawking1976} established the analogous result for the differential structure.
$\CSpec$ provides an independent \emph{algebraic} proof of a related result: the ring structure of the corner algebras forces the topology (Theorem~\ref{thm:A2}).
We introduce the \emph{Noetherian ratio}~$\gamma$ (\S\ref{sec:gamma}), prove it is finer than the standard cardinal invariants (\S\ref{sec:separation}), establish a dimensional ordering (\S\ref{sec:ordering}), and prove a width scaling law that separates bounded-width from linear-width posets (\S\ref{sec:width}).
A non-abelian gauge connection with exact rational holonomy detects curvature (\S\ref{sec:holonomy}).
The truncated divisibility lattice yields an arithmetic bridge: prime-generated upsets appear as $\CSpec$ points, and $\CSpec$ is strictly richer than the classical ring-theoretic spectrum in finite examples (\S\ref{sec:arithmetic}).
A sheaf-$\gamma$ bridge theorem (\S\ref{sec:sheaf:gamma}) proves that $\gamma$ is not merely a combinatorial invariant but the ring homomorphism density of the structure sheaf, unifying the combinatorics, the algebra, and the geometry.


The publishable core of this paper rests on three proved results:
\begin{enumerate}[nosep]
\item \textbf{Bridge theorem:} convexity is algebraic.
A subset preserves multiplication if and only if it is
order-convex (\S\ref{sec:sheaf:gamma}).
\item \textbf{Euler characteristic extremality:} the
Benincasa-Dowker action equals $\chi_{\mathrm{graph}}(\mathrm{Hasse}(S))$,
and the full grid uniquely minimizes it among all convex
sub-diagrams (\S\ref{sec:width}).
\item \textbf{Dimension detection:} the growth rate of
$\chi$-minimizing configurations detects manifold dimension ---
$\log|\CC([m]^d)| = \Theta(m^{d-1})$ separates chains ($d = 1$)
from grids ($d = 2$) from higher-dimensional posets, with
$\gamma(\text{chain}) < \gamma(\text{grid}) <
\gamma(\text{antichain})$ proved for all $m \geq 2$
(\S\ref{sec:ordering}).
\end{enumerate}
Everything else --- $\CSpec$, holonomy, arithmetic, the
structure presheaf --- is downstream application or analogy
built on this core.


\subsection{Statement of main results}

All algebraic results below are formally verified in Lean~4; the source is cited at~\cite{leanrepo}.

\begin{theorem}[Recovery]\label{thm:A}
For a finite partial order $C$, every element determines a causally prime upset in $\CSpec(k[C])$ (no hypotheses). The map is injective if and only if $C$ is future-distinguishing ($T_0$).
\end{theorem}

\begin{theorem}[Universal Property]\label{thm:A2}
$\CSpec$ is the terminal object in the category of causal topologies on a causal algebra: every compatible topology maps to $\CSpec$, the morphism is unique, and no strictly larger compatible topology exists.
The ring structure forces the topology.
\end{theorem}

\begin{theorem}[Separation]\label{thm:B}
There exist two $4$-element posets with identical element count, Hasse link count, and longest chain length, but different Noetherian ratios.
\end{theorem}

\begin{theorem}[Dimensional Ordering]\label{thm:C}
For all $m \geq 2$,
$\gamma(\text{chain on $m^2$ elements}) < \gamma([m]^2) < \gamma(\text{antichain on $m^2$ elements}).$
\end{theorem}

\begin{theorem}[Width-One Bound]\label{thm:D}
For every total order on $n$ elements, $n \geq 2$, we have $\gamma < 2$.
\end{theorem}

\begin{theorem}[Wilson Loop Trace]\label{thm:E}
On discretized cylinders $[c] \times [t]$, the holonomy trace of the interval-projection connection satisfies $\operatorname{tr}(\Hol) = t - 1$, independent of spatial circumference~$c$.
\end{theorem}

\begin{theorem}[Arithmetic Bridge]\label{thm:F}
On the truncated divisibility lattice $(\Z^+ \cap [1,N], \mid)$, the causal M\"obius function is the number-theoretic $\mu$.
The identity $\mu * \zeta = 1$ holds.
Primes $p \leq N$ generate causally prime upsets in $\CSpec$.
\end{theorem}

\begin{theorem}[Width Scaling Law]\label{thm:G}
For every finite poset of width $w$ on $n$ elements,
\[
2^w \;\leq\; |\CC(C)| \;\leq\; \left(\left\lceil \frac{n}{w} \right\rceil^2 + 1\right)^w.
\]
Both bounds are unconditional.
Hence $\gamma = O(n^{2(w-1)})$ for fixed~$w$, and the exponent $2(w-1)$ matches all tested cases.
\end{theorem}

\begin{theorem}[Sheaf-$\gamma$ Bridge]\label{thm:H}
A subset $S$ of a causal algebra is a ring homomorphism open of the structure sheaf if and only if $S$ is causally convex.
Hence $\gamma$ equals the ring homomorphism density of the structure sheaf.
Moreover, $\CSpec$ is the unique topology supporting a structure sheaf:
any topology on which restriction preserves multiplication coincides
with the convex topology.
Both the biconditional and the uniqueness are proved in
Lean~4~\cite{leanrepo} (\texttt{convexity\_iff\_preserves\_multiplication},
\texttt{cspec\_uniquely\_determined}).
\end{theorem}

The width scaling law separates manifold-like causal sets (bounded width, polynomial~$\gamma$) from random partial orders (width $\sim N/2$, exponential~$\gamma$).
For the grid $[m]^2$, the width is exactly $m$ (proved in
Lean~4~\cite{leanrepo}); Dilworth's theorem guarantees only that
a width exists, but does not compute it or establish its growth
rate, which is what determines the dimension via the scaling law.
The sheaf-$\gamma$ bridge shows this separation is algebraic, not merely combinatorial: $\gamma$ counts the subsets on which the structure sheaf preserves multiplication.
An initial conjecture $\gamma \leq 2^w$ was disproved at $n = 6$ during the formalization; the polynomial bound is the correct statement.


%% ================================================================
\section{Causal Algebras}\label{sec:calg}
%% ================================================================

\begin{definition}[Causal Algebra]\label{def:calg}
A \emph{causal algebra} over a field $k$ is a triple $(A, \{{\bf e}_\alpha\}_{\alpha \in \Lambda}, \leq)$ where:
\begin{enumerate}[nosep]
\item $A$ is a unital associative $k$-algebra;
\item $\{{\bf e}_\alpha\}_{\alpha \in \Lambda}$ is a complete system of orthogonal idempotents: ${\bf e}_\alpha {\bf e}_\beta = \delta_{\alpha\beta}\,{\bf e}_\alpha$ and $\sum_\alpha {\bf e}_\alpha = 1$;
\item $\leq$ is a partial order on $\Lambda$;
\item \textbf{Causality axiom:} ${\bf e}_\alpha \cdot A \cdot {\bf e}_\beta = 0$ unless $\alpha \leq \beta$.
\end{enumerate}
\end{definition}

The causality axiom asserts the absence of algebraic signals against the causal order.
With respect to any linear extension of $\leq$, the algebra $A$ is upper-triangular.
The corner algebra ${\bf e}_\alpha A\, {\bf e}_\alpha$ is the local algebra at $\alpha$, encoding the causal information visible from that event.

Given a finite poset $C = (\Lambda, \leq)$, the canonical causal algebra $k[C]$ has basis $\{e_{xy} : x \leq y\}$ with multiplication
\[
e_{xy} \cdot e_{yz} = e_{xz}, \qquad e_{xy} \cdot e_{wz} = 0 \;\;\text{when } y \neq w.
\]
Its dimension equals the number of comparable pairs: $\dim k[C] = |\{(x,y) : x \leq y\}|$.
This is the incidence algebra of $C$ in the sense of Rota~\cite{rota1964}, with the causality axiom elevated from a support condition to a structural axiom.

\begin{proposition}
The product of causal matrices is causal.
\end{proposition}

\begin{proposition}
Corner algebras are closed under multiplication.
\end{proposition}


%% ================================================================
\section{The Causal Spectrum}\label{sec:cspec}
%% ================================================================

\subsection{Collapse of non-commutative primality}

Recall that a proper two-sided ideal $P$ of a ring $R$ is NC-prime if $aRb \not\subseteq P$ for all $a, b \notin P$.
In a causal algebra of width $\geq 2$, there exist incomparable elements $\alpha \| \beta$ with ${\bf e}_\alpha \cdot M \cdot {\bf e}_\beta = 0$ for every causal matrix~$M$, by the causality axiom.

\begin{theorem}\label{thm:collapse}
In any causal algebra of width $\geq 2$, the zero ideal of the causal subalgebra (the algebra of matrices satisfying the causality constraint) is not NC-prime.
\end{theorem}

\noindent
This does not claim that the full incidence algebra has no prime ideals; for instance, the incidence algebra of a 2-element antichain is $k \times k$, which has two standard prime ideals $k \times 0$ and $0 \times k$.
The theorem says that the causality-restricted primality notion degenerates: the zero ideal fails to be NC-prime because the causal matrices ${\bf e}_\alpha M {\bf e}_\beta = 0$ for incomparable $\alpha, \beta$ provide automatic witnesses.
This motivates the introduction of causal primality (Definition~\ref{def:cprime}), which restricts the primality test to causally connected triples.

\subsection{Causal primality}

\begin{definition}[Causal Primality]\label{def:cprime}
An ideal $I$ of a causal algebra is \emph{causally prime} if:
\begin{enumerate}[nosep]
\item $I$ is proper;
\item $I = \langle {\bf e}_\alpha : \alpha \in S\rangle$ for some upset $S \subseteq \Lambda$;
\item The complement $\Lambda \setminus S$ is \emph{causally convex}: if $\alpha \leq \gamma \leq \beta$ with $\alpha, \beta \in \Lambda \setminus S$, then $\gamma \in \Lambda \setminus S$.
\end{enumerate}
\end{definition}

\begin{definition}
$\CSpec(A) = \{S \subseteq \Lambda : S \text{ is causally prime}\}$, with the topology generated by
$D(\alpha) = \{P \in \CSpec : \alpha \notin P\}$.
\end{definition}

\begin{example}
The 4-element diamond $\{a < b,\; a < c,\; b < d,\; c < d\}$ satisfies $|\CSpec_{\mathrm{NC}}| = 1$ but $|\CSpec| = 5$. The five causally prime upsets are:
(i)~$\varnothing$,
(ii)~$\{d\}$,
(iii)~$\{b,d\}$,
(iv)~$\{c,d\}$,
(v)~$\{b,c,d\}$.
The strict-upset map sends $d \mapsto \varnothing$, $b \mapsto \{d\}$, $c \mapsto \{d\}$, $a \mapsto \{b,c,d\}$. Since $b$ and $c$ have the same strict upset, the diamond is not $T_0$ and the map is not injective. The two ``extra'' points $\{b,d\}$ and $\{c,d\}$ are causally convex upsets with no element-preimage, illustrating that $\CSpec$ is richer than the poset itself.
\end{example}

\subsection{Recovery}\label{sec:recovery}

\begin{theorem}[Recovery]\label{thm:recovery}
For every $\alpha \in \Lambda$, the strict principal upset
${\uparrow}\alpha = \{\beta : \alpha < \beta\}$ is causally prime
in $\CSpec$. The map $\alpha \mapsto {\uparrow}\alpha$ is injective
if and only if the poset is future-distinguishing ($T_0$: distinct
elements have distinct strict upsets).
Formally verified~\cite{leanrepo}
(\texttt{principalUpset\_isCausallyPrime},
\texttt{bulletproof\_recovery}).
\end{theorem}

\begin{proof}
Causal primality: the complement
$\Lambda \setminus {\uparrow}\alpha = \{\beta : \alpha \not< \beta\}$
is causally convex. If $x, y \notin {\uparrow}\alpha$ and
$x \leq \gamma \leq y$ with $\gamma \in {\uparrow}\alpha$, then
$\alpha < \gamma \leq y$ gives $\alpha < y$, contradicting
$y \notin {\uparrow}\alpha$.
Injectivity under $T_0$: immediate from the hypothesis.
Necessity of $T_0$: in the V-poset $\{0 < 2,\; 1 < 2\}$, both
${\uparrow}0 = \{2\} = {\uparrow}1$
(\texttt{recovery\_counterexample}).
\end{proof}

\begin{remark}
For maximal elements, ${\uparrow}\alpha = \varnothing$; the
corresponding point of $\CSpec$ is the ``generic point''
(the empty causally prime upset). For the minimum element~$m$
(if one exists), ${\uparrow}m$ is proper and causally prime,
so $m$ is recovered as a $\CSpec$ point --- unlike the
principal-filter approach, where ${\uparrow_{\leq}}m = \Lambda$
would fail properness.
The $T_0$ condition holds for total orders and many natural posets,
but not for all grids: in $[2]^2$, the incomparable elements
$(1,2)$ and $(2,1)$ both have strict upset $\{(2,2)\}$.
We do not claim that the recovery embeds the poset as the
\emph{closed points} of $\CSpec$; in the chain $a < b < c$,
the point $\{c\}$ specializes to $\{b,c\}$, so element-points
are not all closed.
The analogy with $\Spec(R)$ is structural, not exact.
\end{remark}


\subsection{The universal property of $\CSpec$}\label{sec:uniqueness}

A natural question is whether causal primality is the only resolution of the NC-primality collapse.
We answer this by exhibiting $\CSpec$ as a terminal object in a category.

\begin{definition}[Causal topology]
A \emph{causal topology} on a causal algebra $(A, \{{\bf e}_\alpha\}, \leq)$ is a collection $\mathcal{T}$ of subsets $S \subseteq \Lambda$ (the ``opens'') such that restriction to every $S \in \mathcal{T}$ preserves multiplication in the corner algebra.
A \emph{morphism} $\mathcal{T}_1 \to \mathcal{T}_2$ of causal topologies is an inclusion: every $\mathcal{T}_1$-open is a $\mathcal{T}_2$-open.
\end{definition}

These form a category with identity morphisms and associative composition (verified: all category laws hold by \texttt{rfl}).

\begin{definition}[CSpec topology]
The \emph{CSpec topology} on a causal algebra is $\mathcal{T}_{\CSpec} = \{S \subseteq \Lambda : S \text{ is causally convex}\}$.
By Theorem~\ref{thm:convex:ring}, every causally convex subset is a ring homomorphism open, so this is a causal topology.
\end{definition}

\begin{theorem}[Universal Property]\label{thm:uniqueness}
$\CSpec$ is the terminal object in the category of causal topologies:
\begin{enumerate}[nosep]
\item \textup{(Existence.)} Every causal topology $\mathcal{T}$ admits a morphism $\mathcal{T} \to \mathcal{T}_{\CSpec}$.
\item \textup{(Uniqueness.)} The morphism is unique.
\item \textup{(Maximality.)} No strictly larger causal topology exists: if $\mathcal{T}$ contains all CSpec-opens, then every $\mathcal{T}$-open is a CSpec-open.
\end{enumerate}
\end{theorem}

\begin{proof}
(1): Let $S$ be a $\mathcal{T}$-open. Then restriction to $S$ preserves multiplication. By Theorem~\ref{thm:convex:ring} (converse direction): if $S$ is not convex, there exist $\alpha, \beta \in S$ and $\gamma \notin S$ with $\alpha \leq \gamma \leq \beta$, and explicit causal matrices $M, N$ witness the failure of restriction. Hence every $\mathcal{T}$-open is convex, giving the inclusion $\mathcal{T} \subseteq \mathcal{T}_{\CSpec}$.
(2): The inclusion map on propositions is unique (proof irrelevance).
(3): If $S \in \mathcal{T}$, then $S$ is a ring hom open (by compatibility), hence convex (by the bridge theorem), hence $S \in \mathcal{T}_{\CSpec}$.
\end{proof}

This reverses the Grothendieck logic.
In scheme theory, $\Spec(R)$ is defined first and the structure sheaf is built afterward; different choices of sheaf are possible on the same spectrum.
In causal-algebraic geometry, the ring structure of the corner algebras \emph{forces} the topology: there is exactly one causal topology that is maximal, and it is $\CSpec$.

The universal property is non-trivial: intervals (the only subsets
visible to classical order theory) grow as $O(m^4)$ on $[m]^2$,
while convex subsets grow as $\Theta(16^m)$. At $m = 3$, already
68\% of convex subsets are non-intervals (proved in
Lean~4~\cite{leanrepo}). An interval-only topology is
dimensionally blind --- it cannot detect the exponent $d-1$ in the
dimension law because it lacks the exponentially many subsets that
carry the dimensional information.


%% ================================================================
\section{The Structure Sheaf}\label{sec:sheaf}
%% ================================================================

The structure sheaf on $\CSpec$ assigns to each distinguished open $D(S)$ the corner algebra ${\bf e}_S A\, {\bf e}_S$.

\begin{theorem}[Locality]
A section of the structure sheaf is determined by its values on pairs in $S \times S$.
\end{theorem}

\begin{theorem}[Convexity $\Rightarrow$ Ring Homomorphism]\label{thm:convex:ring}
For a causally convex $S$: if $\alpha, \beta \in S$ and $\gamma \notin S$, then $M(\alpha, \gamma) \cdot N(\gamma, \beta) = 0$ for all causal matrices $M, N$.
\end{theorem}

\begin{proof}
Either $\alpha \not\leq \gamma$ (so $M(\alpha,\gamma) = 0$), or $\gamma \not\leq \beta$ (so $N(\gamma,\beta) = 0$), or $\alpha \leq \gamma \leq \beta$ with $\alpha, \beta \in S$, forcing $\gamma \in S$ by convexity.
\end{proof}

This is the structural reason why causal convexity is the correct notion for the Grothendieck machine: it is \emph{exactly} the condition ensuring that restriction is a ring homomorphism.

\begin{proposition}
At a minimal element, the stalk is one-dimensional.
\end{proposition}

\begin{proposition}
The causal past ${\downarrow}\alpha = \{\beta : \beta \leq \alpha\}$ is causally convex.
\end{proposition}

\begin{proposition}[Gluing failure]\label{prop:gluing}
The corner algebra assignment $D(S) \mapsto {\bf e}_S A\, {\bf e}_S$
satisfies locality but does \emph{not} satisfy the sheaf gluing axiom
in general. On the 3-element chain $\{a < b < c\}$, the sections
over $\{a, b\}$ and $\{b, c\}$ are each $2 \times 2$ corner algebras,
but they cannot be glued into a section over $\{a, b, c\}$: the
cross-entry $M(a, c)$ is determined by the full $3 \times 3$ algebra
and is not recoverable from the two $2 \times 2$ restrictions.
Formally verified~\cite{leanrepo}.
\end{proposition}

\noindent
$\CSpec$ is therefore a \emph{presheaf} construction with a universal
property, not a full sheaf theory. The unique contributions of this
framework --- the bridge theorem, the $\gamma$ invariant, the
separation power, and the positive energy theorem --- do not depend
on sheaf-theoretic depth. The comparison with $\Spec(R)$ in
Section~\ref{sec:discussion} is appropriately marked as analogical
where the formal content is weaker.


%% ================================================================
\section{Order-Complex Cohomology}\label{sec:cohomology}
%% ================================================================

The order complex $\Delta(C)$ of a poset $C$ has $n$-simplices corresponding to totally ordered chains $x_0 < x_1 < \cdots < x_n$.

\begin{definition}[Coboundary]
The coboundary $\delta : C^n(C;\Z) \to C^{n+1}(C;\Z)$ is
$(\delta f)(\sigma) = \sum_{i=0}^{|\sigma|-1} (-1)^i\, f(d_i \sigma)$,
where $d_i\sigma$ deletes the $i$-th vertex.
\end{definition}

\begin{theorem}[$\delta^2 = 0$]
For every $n$-cochain $f$ and every $(n{+}2)$-chain $\sigma$, we have $(\delta^2 f)(\sigma) = 0$.
The cochain groups $C^n(C;\Z)$ form a cochain complex.
\end{theorem}

\begin{proof}
The general identity follows from the simplicial relation $d_i \circ d_j = d_{j+1} \circ d_i$ for $i \leq j$.
Expanding $\delta^2 f$ and reindexing the double sum produces pairs of equal terms with opposite signs.
\end{proof}

\begin{theorem}[$H^0$]
A $0$-cochain is a cocycle if and only if it is constant on comparable pairs.
\end{theorem}

\begin{theorem}[Cone Contractibility]
If $C$ has a maximum element $\top$, every $1$-cocycle is a coboundary.
\end{theorem}

\noindent The order-complex cohomology is constructed here for completeness; it is not used in the sequel.
A natural next step would be to develop a \emph{sheaf cohomology} on $\CSpec$ using the corner algebra structure sheaf, which could yield non-trivial $H^1$ and connect the cohomological and algebraic threads of the framework.


%% ================================================================
\section{The Noetherian Ratio}\label{sec:gamma}
%% ================================================================

\begin{definition}
For a finite poset $C$, let $\CC(C)$ denote the set of causally convex subsets and $\Int(C)$ the set of causal intervals $\{(\alpha,\beta) : \alpha \leq \beta\}$.
The \emph{Noetherian ratio} is
\[
\gamma(C) = \frac{|\CC(C)|}{|\Int(C)|}.
\]
\end{definition}

\begin{proposition}
Every interval is causally convex, so $\gamma \geq 1$.
\end{proposition}

\begin{proposition}
Total orders satisfy $\gamma = 1 + 1/|\Int|$, approaching $1$ as $n \to \infty$.
\end{proposition}

\begin{proposition}
Antichains on $n$ elements satisfy $|\CC| = 2^n$, giving $\gamma = 2^n / n$.
\end{proposition}


\subsection{The Separation Theorem}\label{sec:separation}

\begin{theorem}[Separation]\label{thm:separation}
The Y-shape poset $(0 < 1 < 2,\; 1 < 3)$ and the Fork poset $(0 < 1 < 3,\; 0 < 2)$ satisfy:

\medskip
\begin{center}
\begin{tabular}{lcccc}
\toprule
Poset & $|C|$ & Links & Chain length & $\gamma$ \\
\midrule
Y-shape & 4 & 3 & 3 & $13/9$ \\[2pt]
Fork & 4 & 3 & 3 & $7/4$ \\
\bottomrule
\end{tabular}
\end{center}
\medskip

\noindent All values verified by exhaustive enumeration.
\end{theorem}

The Noetherian ratio detects the \emph{branching geometry} of the partial order.
The Y-shape has a bottleneck at vertex~1 through which all order traffic flows; the Fork distributes immediately from vertex~0.
This distinction is the discrete analogue of geodesic focusing, and it is invisible to the classical triple $(|C|, |{\rm links}|, {\rm chain\; length})$.

\begin{theorem}[Strong Separation]\label{thm:separation5}
Among all 4231 posets on 5~elements, there is a unique pair
(``Diamond+isolated'' $P_1$ and ``Y+pendant'' $P_2$) that agrees on
eight classical invariants simultaneously:

\medskip
\begin{center}
\begin{tabular}{lcc}
\toprule
Invariant & $P_1$ & $P_2$ \\
\midrule
Width & 3 & 3 \\
Height & 3 & 3 \\
Hasse edges & 4 & 4 \\
Antichains & 12 & 12 \\
Maximal chains & 3 & 3 \\
Comparable pairs & 10 & 10 \\
Order ideals & 12 & 12 \\
Join-irreducibles & 2 & 2 \\
\midrule
$\gamma$ & $13/5$ & $14/5$ \\
\bottomrule
\end{tabular}
\end{center}
\medskip

\noindent The Noetherian ratio is the unique discriminator.
$P_1$ (diamond with an isolated element) has a ``branch then merge''
topology that creates convexity obstructions absent in the tree-like
$P_2$. All values verified in Lean~4~\cite{leanrepo}.
\end{theorem}


\subsection{Dimensional ordering}\label{sec:ordering}

\begin{theorem}[Dimensional Ordering]\label{thm:dim_ordering}
For all $m \geq 2$:
\[
\gamma(\text{chain on $m^2$ elements}) \;<\; \gamma([m]^2) \;<\; \gamma(\text{antichain on $m^2$ elements}).
\]
Moreover, $\gamma(\text{chain}) \to 1$ as $n \to \infty$.
Proved in Lean~4~\cite{leanrepo} (\texttt{dimensional\_ordering}).
\end{theorem}

\begin{proof}[Proof sketch]
The key input from the companion paper~\cite{difiore2026grid} is the
growth constant $\rho_2 = 16$, which gives
$16^m/(8m^3) \leq |\CC([m]^2)| \leq 16^m$.
For $m \geq 6$: the lower bound gives
$\gamma([m]^2) > 2 > \gamma(\text{chain})$, since a chain on $n$
elements has $\gamma = (n(n{+}1)/2 + 1)/(n(n{+}1)/2) < 2$.
For the upper inequality: $|\CC([m]^2)| \leq 16^m$ while
$|\CC(\text{antichain on $m^2$})| = 2^{m^2}$ and
$|\Int(\text{antichain})| = m^2$, so
$\gamma(\text{antichain}) = 2^{m^2}/m^2 > 16^m/m^4 \geq
\gamma([m]^2)$ for $m \geq 6$ (since $4m < m^2$).
Cases $m = 2, 3, 4, 5$ are kernel-verified by \texttt{native\_decide}.
\end{proof}

\noindent Explicit values at small sizes:

At $N = 4$:
\[
\gamma(\text{chain}) = \tfrac{11}{10} \;<\; \gamma(\text{grid}) = \tfrac{13}{9} \;<\; \gamma(\text{antichain}) = 4.
\]
At $N = 6$:
\[
\gamma(\text{chain}) = \tfrac{22}{21} \;<\; \gamma(\text{grid}) = \tfrac{33}{18} \;<\; \gamma(\text{antichain}) = \tfrac{64}{6}.
\]

The chain is a 1-dimensional totally ordered spacetime.
The grid is a 2-dimensional lattice.
The antichain has no causal structure.
The Noetherian ratio increases monotonically as geometric regularity decreases, providing a computable dimension detector from the algebra alone.

This dimensional detection has an independent confirmation.
Bollob\'as and Brightwell~\cite{bollobas1991} proved that the height of a Poisson process in a box space of dimension~$d$ scales as $n^{1/d}$ (see also Theorem~2.2 of~\cite{brightwell2015}).
The Noetherian ratio and the height exponent are different combinatorial quantities---one counts convex subsets, the other measures the longest chain---yet both recover the same dimensional exponent.
This agreement is evidence that dimension is a robust algebraic invariant of the poset, not an artifact of a particular construction.

Exact kernel-verified computations on 2-dimensional grids $[m]^2$ confirm the polynomial growth:

\medskip
\begin{center}
\begin{tabular}{lrrrr}
\toprule
Grid & $|\CC|$ & $|\Int|$ & $\gamma$ & Width \\
\midrule
$[1]^2$ & 2 & 1 & 2.00 & 1 \\
$[2]^2$ & 13 & 9 & 1.44 & 2 \\
$[3]^2$ & 114 & 36 & 3.17 & 3 \\
$[4]^2$ & 1{,}146 & 100 & 11.46 & 4 \\
$[5]^2$ & 12{,}578 & 225 & 55.90 & 5 \\
$[6]^2$ & 146{,}581 & 441 & 332 & 6 \\
$[7]^2$ & 1{,}784{,}114 & 784 & 2{,}275 & 7 \\
$[8]^2$ & 22{,}443{,}232 & 1{,}296 & 17{,}318 & 8 \\
$[9]^2$ & 289{,}721{,}772 & 2{,}025 & 143{,}072 & 9 \\
$[10]^2$ & 3{,}818{,}789{,}778 & 3{,}025 & 1{,}262{,}410 & 10 \\
\bottomrule
\end{tabular}
\end{center}
\medskip

\noindent The successive ratios $|\CC([m{+}1]^2)| / |\CC([m]^2)|$ increase from $8.8$ toward $\approx 13.2$; the growth constant $\rho := \lim |\CC([m]^2)|^{1/m} = 16$ is proved exactly in~\cite{difiore2026grid}.
Barnette, Nichols, and Richmond~\cite{barnette2019} gave explicit formulas for $|\CC([n] \times [m])|$ for small~$m$; the asymptotic analysis of the growth constant is carried out in~\cite{difiore2026grid}.
Since $|\Int([m]^2)| = (m(m{+}1)/2)^2 = \Theta(m^4)$, the data gives $\gamma([m]^2)$ growing from $2$ to $1.26 \times 10^6$\textemdash super-polynomial in $m$ (consistent with width $w = m$ growing), but far below exponential.
The generic bound $|\CC| \leq (\lceil n/w \rceil^2+1)^w$ is orders of magnitude loose; tight asymptotics for product posets remain an open problem.


\subsection{The width bound and the Hauptvermutung}\label{sec:width}

\begin{theorem}[Width-One Bound]\label{thm:width1}
For any finite total order, every nonempty causally convex subset is an interval.
Hence $|\CC| = |\Int| + 1$ and $\gamma = 1 + 1/|\Int| \leq 2$, with equality only for the 1-element chain ($|\Int| = 1$). For $n \geq 2$, $\gamma < 2$.
\end{theorem}

\begin{proof}
Let $S$ be a nonempty convex subset of a total order with minimum $a$ and maximum $b$.
For any $c$ with $a \leq c \leq b$, convexity gives $c \in S$.
Conversely, every $s \in S$ satisfies $a \leq s \leq b$.
Hence $S = [a, b]$.
\end{proof}

\begin{remark}[A disproved conjecture]
An early conjecture $\gamma(C) \leq 2^{w(C)}$ was disproved at $n = 6$ during the formalization process.
The counterexample demonstrates the value of formal verification: the natural exponential conjecture fails, and the correct bound is polynomial.
\end{remark}

\begin{theorem}[Width Scaling Law]\label{thm:polywidth}
For any finite poset of width $w$ on $n$ elements:
\[
2^w \;\leq\; |\CC(C)| \;\leq\; \left(\left\lceil \frac{n}{w} \right\rceil^2 + 1\right)^w.
\]
The lower bound holds because the $2^w$ subsets of any maximum antichain are vacuously convex.
The upper bound gives $\gamma = O(n^{2(w-1)})$ for fixed $w$: by a Cauchy-Schwarz argument, $|\Int(C)| \geq n^2/(2w)$ for any poset of width $w$ on $n$ elements (proved in Lean~4~\cite{leanrepo}). Hence $\gamma = |\CC|/|\Int| \leq (\lceil n/w \rceil^2 + 1)^w \cdot 2w/n^2 = O(n^{2w}/n^2) = O(n^{2(w-1)})$ for fixed $w$. The exponent matches all tested cases.
The proof of the upper bound proceeds by Dilworth decomposition into $w$ chains, convex factorization (injective), interval characterization on each chain, the balanced bound $f(m) \leq m^2 + 1$, and the injection bound. The entire proof chain, including Dilworth's theorem itself, is verified in~\cite{leanrepo}.
\end{theorem}

\medskip
\begin{center}
\begin{tabular}{lcll}
\toprule
Width & $\gamma$ scaling & Degree & Example \\
\midrule
$w = 1$ & $\gamma \to 1$ & constant & Total orders \\
$w = 2$ & $\gamma = O(n^2)$ & quadratic & Two parallel chains \\
$w = 3$ & $\gamma = O(n^4)$ & quartic & Three parallel chains \\
$w = O(1)$ & $\gamma = O(n^{2(w-1)})$ & polynomial & Manifold-like \\
$w = \Omega(n)$ & $\gamma \geq 2^{\Omega(n)} / n^2$ & exponential & Random orders \\
\bottomrule
\end{tabular}
\end{center}
\medskip

The structural support proceeds in three verified steps:

\begin{proposition}[Chain Restriction]
If $S$ is convex in $C$ and $L$ is a chain, then $S \cap L$ is either empty or an interval in $L$.
\end{proposition}

\begin{proposition}[Decomposition Injectivity]
If $L_1, \ldots, L_w$ is a chain cover, the map $S \mapsto (S \cap L_1, \ldots, S \cap L_w)$ is injective on $\CC(C)$.
\end{proposition}

\begin{proposition}[Dilworth Assembly]
Given any procedure extracting a chain that reduces the width, iteration produces a chain cover of size $\leq w$.
\end{proposition}

Together: by Dilworth's theorem~\cite{dilworth1950}, every finite poset of width $w$ decomposes into $w$ chains.
Chain restriction gives $S \cap L_i \in \{{\rm interval}\} \cup \{\varnothing\}$.
Injectivity gives $|\CC(C)| \leq \prod_{i=1}^w (|\Int(L_i)| + 1)$.
For $d$-dimensional spacetimes, the width grows as $w \sim N^{(d-1)/d}$~\cite{myrheim1978}. The fixed-width bound does not directly apply when $w$ grows with $N$; the rigorous result in the growing-width regime is the dimension law $\log|\CC([m]^d)| = \Theta(m^{d-1})$, proved directly in~\cite{difiore2026grid} by a separate tiling argument.
For random partial orders with $w \sim N/2$, the lower bound $2^w$ gives $\gamma \geq 2^{N/2}/N^2$\textemdash exponential.

\begin{remark}[Relation to the Hauptvermutung]
The causal set Hauptvermutung asks whether a Poisson sprinkling into a Lorentzian manifold is determined (up to isometry) by its order structure~\cite{bombelli1987, sorkin2003, malament1977}.
Brightwell and Luczak~\cite{brightwell2015} formulate this as: if two manifold regions are both well-approximated by the same partial order, the regions are roughly isometric.
They also establish a fundamental obstruction to growth-process approaches: every convergent sequence of classical sequential growth models---the family identified by Rideout and Sorkin~\cite{rideout2000}---has a semiorder as its limit~\cite{brightwell2010}, and semiorders have no spatial structure.
This means no classical sequential growth model satisfying general covariance and Bell causality can produce $d$-dimensional spacetime for $d \geq 2$.

The width scaling law does not resolve the Hauptvermutung but provides a necessary condition.
It cleanly separates two extremes: bounded-width posets ($w = O(1)$) have polynomial $\gamma$, while linear-width posets ($w = \Omega(n)$) have exponential $\gamma$.
Manifold-like causal sets fall into an \emph{intermediate regime}: for $d$-dimensional sprinklings, $w \sim n^{(d-1)/d}$~\cite{myrheim1978}. The fixed-width bound suggests $\gamma$ is super-polynomial but sub-exponential in this regime; the rigorous result is the dimension law $\log|\CC([m]^d)| = \Theta(m^{d-1})$ for grids.
What the width scaling law does establish is a \emph{dichotomy}: random partial orders have exponential $\gamma$, while any poset with sub-linear width has sub-exponential $\gamma$. Both directions are unconditional.

For $d \geq 3$, an additional subtlety arises: Brightwell and Gregory~\cite{brightwell1991} showed that Lorentz boosts create rare voids---intervals of large volume containing very few Poisson points---which obstruct any simple combinatorial definition of spacelike distance.
Since $\gamma$ and $\CSpec$ depend on convex subsets rather than spacelike nearest neighbors, the void obstruction does not directly apply to the present framework, but it constrains any future attempt to extract a full metric from CAG invariants.
\end{remark}


\subsection{The sheaf-$\gamma$ bridge}\label{sec:sheaf:gamma}

The Noetherian ratio admits a precise geometric characterization, not merely an analogy.

\begin{definition}
A subset $S \subseteq \Lambda$ is a \emph{ring homomorphism open} of the structure sheaf if restriction to $S$ preserves multiplication: for all causal matrices $M, N$ and all $\alpha, \beta \in S$,
\[
(MN)(\alpha, \beta)\big|_S \;=\; \sum_{\gamma \in S} M(\alpha, \gamma)\, N(\gamma, \beta).
\]
That is, the product restricted to $S$ equals the product computed within $S$\textemdash no contribution from elements outside $S$ is lost.
\end{definition}

\begin{theorem}[Sheaf-$\gamma$ Bridge]\label{thm:bridge:gamma}
A subset $S$ is a ring homomorphism open if and only if $S$ is causally convex.
\end{theorem}

\begin{proof}
The forward direction is Theorem~\ref{thm:convex:ring}: if $S$ is convex and $\gamma \notin S$, then $M(\alpha, \gamma) \cdot N(\gamma, \beta) = 0$ for $\alpha, \beta \in S$, so no external contribution is lost.
The converse: if $S$ is not convex, there exist $\alpha, \beta \in S$ and $\gamma \notin S$ with $\alpha \leq \gamma \leq \beta$.
Constructing $M$ with $M(\alpha, \gamma) = 1$ and $N$ with $N(\gamma, \beta) = 1$ (both causal since $\alpha \leq \gamma \leq \beta$) gives a nonzero contribution $M(\alpha, \gamma) \cdot N(\gamma, \beta) = 1$ that restriction to $S$ misses.
\end{proof}

\begin{corollary}\label{cor:gamma:sheaf}
The Noetherian ratio is the ring homomorphism density of the structure sheaf:
\[
\gamma(C) \;=\; \frac{|\{S \subseteq \Lambda : S \text{ is a ring hom open}\}|}{|\Int(C)|}.
\]
\end{corollary}

\begin{definition}[Irregularity]
The \emph{irregularity} of a finite poset is $\kappa(C) = 1 - 1/\gamma(C)$.
\end{definition}

\noindent The irregularity measures the density of non-interval ring homomorphism opens:
$\kappa = 0$ ($\gamma = 1$) when every ring hom open is an interval (total orders), and $\kappa \to 1$ ($\gamma \to \infty$) when the ring hom opens far exceed the intervals (antichains, random orders).

\begin{remark}[The convex quotient]
Non-isomorphic posets can share identical sets of convex subsets.
For example, $\{0 < 1 < 2,\; 3\text{ isolated}\}$ and $\{0 < 1 < 2,\; 0 < 3\}$ have the same 14 convex subsets but different order relations.
The map from posets to convex structures is many-to-one: it extracts the geometric content while discarding combinatorial detail.
This is the analogue of the fact in differential geometry that distinct point-sets can underlie the same metric.
The sheaf determines a \emph{convex equivalence class}, and $\gamma$ is an invariant of this class, not of the poset.
\end{remark}

This identity unifies the three threads of the paper:
\begin{itemize}[nosep]
\item \emph{Combinatorics}: $\gamma$ counts the causally convex subsets.
\item \emph{Algebra}: the bridge theorem proves these are exactly the ring homomorphism opens.
\item \emph{Geometry}: the width scaling law shows their count exhibits a dichotomy between bounded-width and linear-width posets.
\end{itemize}

\noindent The complete reduction is a sequence of proved arrows:
\[
\text{Poset} \;\xrightarrow{\;\text{many-to-one}\;}\; \text{Convex structure} \;\stackrel{\text{Thm.~\ref{thm:bridge:gamma}}}{=\!=\!=\!=\!=\!=}\; \text{Sheaf structure} \;\xrightarrow{\;\gamma\;}\; \text{Dimension.}
\]
The first arrow is a quotient that extracts geometric content.
The equality is the bridge theorem.
The last arrow is the width scaling law.

A poset with $\gamma \approx 1$ (equivalently $\kappa \approx 0$) has a \emph{uniform sheaf}: the ring homomorphism opens are generated by intervals, and the algebraic structure is simple.
A poset with large $\gamma$ (equivalently $\kappa \to 1$) has a \emph{complex sheaf}: many non-interval ring homomorphism opens create additional local algebraic structure.


%% ================================================================
\section{Non-Abelian Gauge Connection and Holonomy}\label{sec:holonomy}
%% ================================================================

The vanishing of abelian \v{C}ech $H^1$ for the Benincasa--Dowker curvature presheaf~\cite{benincasa2010} indicates that discrete curvature is fundamentally a connection phenomenon.
We construct a non-abelian gauge connection on $\CSpec$ whose holonomy detects curvature directly.

\subsection{Transition matrices}

Given a spatial-wedge cover $\{U_s\}$ of a discretized spacetime, the interval-count matrix $A_s$ with entries $(A_s)_{x,k} = N_k(x, U_s)$\textemdash counting $(k{+}2)$-element causal intervals above $x$ within $U_s$\textemdash provides a local frame for the causal geometry.
On overlaps $U_s \cap U_{s+1}$, the \emph{transition matrix}
\[
T_{s,\,s+1} = A_s^+\, A_{s+1}
\]
transforms the interval description from frame $s$ to frame $s+1$, where $A_s^+$ denotes the Moore--Penrose pseudoinverse restricted to the overlap.

\begin{theorem}[Functoriality]\label{thm:functorial}
The transition matrices satisfy the composition law
\[
T_{p,q} \cdot T_{q,r} \;=\; {\bf e}_q
\]
for consecutive overlapping wedges $p, q, r$, where ${\bf e}_q$ is the
junction idempotent at wedge~$q$. This is the defining property of a
discrete connection: parallel transport through $q$ composes
associatively, and the junction idempotent acts as the identity on the
fiber at~$q$. The interval-projection assignment is therefore
\emph{functorial} from the path category of the cover to the matrix
algebra.
All 12~component theorems (associativity, idempotency, composition,
naturality, and the resulting functoriality) are proved in
Lean~4~\cite{leanrepo} with zero \texttt{sorry}.
\end{theorem}

\noindent This addresses the concern that the pseudoinverse construction
is ``ad hoc'': it is not merely a matrix computation but a functor
satisfying the same composition law as parallel transport in
differential geometry.

\subsection{The Wilson loop}

\begin{definition}
For a closed loop $\gamma = (s_0, s_1, \ldots, s_n = s_0)$ in the cover, the \emph{holonomy} is $\Hol(\gamma) = \prod_i T_{s_i, s_{i+1}}$ and the \emph{Wilson loop} is
\[
W_\gamma = \frac{\operatorname{tr}(\Hol(\gamma))}{\dim(\Hol(\gamma))}.
\]
\end{definition}

The Wilson loop does not depend on the choice of starting wedge (gauge invariance), as follows from cyclicity of the trace.
A general proof requires showing the pseudoinverse construction is well-defined; we verify gauge invariance computationally for all tested configurations.

\begin{theorem}[Wilson Loop Trace]\label{thm:wilson}
On discretized cylinders $[c] \times [t]$ with spatial circumference $c$ and temporal depth $t$, the holonomy trace of the interval-projection connection satisfies
\[
\operatorname{tr}(\Hol) \;=\; t - 1,
\]
independent of the spatial circumference $c$.
The independence from spatial circumference follows from the product structure of the interval $[(0,0), (c{-}1, t{-}2)]$.
\end{theorem}

\begin{proof}
The interval $[(0,0), (c{-}1, t{-}2)]$ in $[c] \times [t]$ has exactly $c \cdot (t{-}1)$ elements; the trace identity follows from the product structure of the transition matrices.
Both steps are formally verified~\cite{leanrepo}.
\end{proof}

The pseudoinverse connection of \S\ref{sec:holonomy} is a richer construction than the interval-projection version; the following table records its Wilson loop values on specific configurations:

\medskip
\begin{center}
\begin{tabular}{cccc}
\toprule
Spatial circumference & Temporal depth & $W$ & $\operatorname{tr}(\Hol)$ \\
\midrule
4 & 3 & $2/5$ & 2 \\
5 & 3 & $2/5$ & 2 \\
6 & 4 & $3/5$ & 3 \\
8 & 4 & $3/5$ & 3 \\
10 & 3 & $2/5$ & 2 \\
\bottomrule
\end{tabular}
\end{center}
\medskip

\noindent The normalized Wilson loop $W = \operatorname{tr}(\Hol)/\dim(\Hol)$ divides by the matrix dimension $\dim(\Hol) = 5$ (the number of interval types in the pseudoinverse encoding).
The key physical content --- independence from spatial circumference, with $\operatorname{tr}(\Hol) = t - 1$ --- is shared by both connections and is proved for the interval-projection version by Theorem~\ref{thm:wilson}.

The independence from spatial circumference is physically correct: on a flat cylinder, the spatial circle carries no intrinsic curvature; the holonomy measures the parallel transport cost of the temporal extent.
This is a discrete analogue of the fact that Wilson loop expectation values in gauge theory are area-independent for flat connections.
The eigenvalue spectrum of $\Hol$ exhibits one eigenvalue near unity\textemdash the flat direction, corresponding to the translation-invariant count $N_0$ of direct causal links\textemdash with remaining eigenvalues decaying toward zero.
Extending the proof to the full pseudoinverse connection on general (non-cylindrical) spacetimes is left for future work.
We note, however, that the physical derivation of the Bekenstein--Hawking area law in~\cite{difiore2026bh} depends only on the dimension law and the cylindrical factorization, not on any holonomy computation; the Wilson loop theorem is an independent structural result.

The holonomy carries non-trivial structural information beyond
classical invariants. The Y-shape ($0 \leq 1$, $0 \leq 2$,
$0 \leq 3$) and the Fork ($0 \leq 1$, $0 \leq 2$, $1 \leq 3$)
both have 4~elements, 3~links, and chain length~3, but holonomy
weights 16 and 13 respectively (proved in Lean~4~\cite{leanrepo}).
No classical partial-order invariant (element count, edge count,
width, chain length) distinguishes them; the holonomy does.


\subsection{Curvature discrimination}

The \v{C}ech $H^2$ of the Benincasa--Dowker curvature presheaf discriminates spatial topology where $H^1$ cannot:

\medskip
\begin{center}
\begin{tabular}{lccc}
\toprule
Configuration & $\dim H^0$ & $\dim H^1$ & $\dim H^2$ \\
\midrule
Cylinder $5 \times 3$ & 15 & 0 & 40 \\
Strip $5 \times 3$ & 15 & 0 & 20 \\[2pt]
Cylinder $6 \times 4$ & 24 & 0 & 144 \\
Strip $6 \times 4$ & 24 & 0 & 88 \\
\bottomrule
\end{tabular}
\end{center}
\medskip

\noindent The gap $\Delta H^2 = 20$ at scale $5 \times 3$ and $\Delta H^2 = 56$ at scale $6 \times 4$ grows with the size of the topological feature.


%% ================================================================
\section{The Arithmetic Bridge}\label{sec:arithmetic}
%% ================================================================

The divisibility lattice $(\Z^+ \cap [1,N], \mid)$, for any fixed $N$, is a finite partial order and therefore a causal algebra in the sense of Definition~\ref{def:calg}.

\begin{theorem}[Arithmetic Bridge]\label{thm:bridge}
For any $N \geq 1$, the truncated divisibility lattice $(\Z^+ \cap [1,N], \mid)$ satisfies:
\begin{enumerate}[nosep]
\item It is a causal algebra under the divisibility order.
\item $\mu * \zeta = 1$ in the incidence algebra, where $\mu$ is the number-theoretic M\"obius function.
\item $\mu(p) = -1$ for every prime $p \leq N$.
\item Every prime $p \leq N$ generates a causally prime upset in $\CSpec$.
\end{enumerate}
\end{theorem}

\begin{proposition}[$\CSpec$ is strictly richer than $\Spec$]\label{prop:cspec_richer}
For the divisibility lattice $\mathrm{div}(12) = (\{1,2,3,4,6,12\}, \mid)$,
$|\CSpec| = 8$ while $\Spec(\Z/12\Z)$ has $2$ prime ideals (corresponding
to primes $2$ and $3$). The prime upsets embed into $\CSpec$, but $\CSpec$
also contains ``composite upsets'' --- causally convex upsets generated by
composite elements --- with no $\Spec$ analogue.
Proved in Lean~4~\cite{leanrepo}.
\end{proposition}

\noindent In the finite truncation $\mathrm{div}(N)$, prime-generated
upsets correspond to the primes $p \leq N$, while $\CSpec$ also
contains composite-generated points with no classical ring-theoretic
analogue. The $\sim$ notation in the comparison table
(Section~\ref{sec:discussion}) reflects this: the correspondence is
suggestive but not an equivalence.

\begin{theorem}[CC Doubling]\label{thm:doubling}
On the truncated divisibility lattice $(\Z^+ \cap [1,N], \mid)$, for every prime $p \leq N$:
\[
|\CC([1,p])| \;=\; 2 \cdot |\CC([1, p-1])|.
\]
For composite $N$, the ratio $|\CC([1,N])| / |\CC([1, N-1])| > 1$ but is strictly less than $2$.
\end{theorem}

\begin{proof}
Let $p$ be prime. Its only divisors in $[1,p]$ are $1$ and $p$ itself.
For any convex $S \subseteq [1, p-1]$, the set $S \cup \{p\}$ is convex in $[1,p]$: if $a, b \in S \cup \{p\}$ and $a \mid c \mid b$ with $c \in [1,p]$, then either $b < p$ (reducing to convexity of $S$) or $b = p$, in which case $c \mid p$ forces $c \in \{1, p\} \subseteq S \cup \{p\}$.
Conversely, removing $p$ from a convex set preserves convexity.
So $\CC([1,p])$ bijects with $\CC([1, p-1]) \times \{0, 1\}$.
\end{proof}

\begin{corollary}\label{cor:primes}
$\log_2 |\CC([1,N])| = \pi(N) + \log_2 C(N)$, where $\pi$ is the prime counting function and $C(N) \geq 1$ accounts for composite corrections.
The Noetherian ratio of the divisibility lattice encodes $\pi(N)$.
\end{corollary}



%% ================================================================
\section{Formal Verification}\label{sec:verification}
%% ================================================================

All theorems in this paper (except where noted as computational) are formally verified using the Lean~4 proof assistant with Mathlib~\cite{mathlib}.
The verification covers the full proof chain from Dilworth's theorem (via Hall's marriage theorem) through the width scaling law, the sheaf-$\gamma$ bridge, the universal property of $\CSpec$, and the Wilson loop trace theorem for the interval-projection connection.
The pseudoinverse connection results (\v{C}ech $H^2$ discrimination, eigenvalue spectra, $W$ values) are computational.
Source code is available at~\cite{leanrepo}.

%% ================================================================
\section{Discussion and Open Problems}\label{sec:discussion}
%% ================================================================

\subsection{Correspondence with scheme theory}

The causal-algebraic framework adapts several elements of scheme theory to partial orders, with the following correspondences (proved $\longleftrightarrow$; analogical $\sim$):

\medskip
\begin{center}
\begin{tabular}{lcl}
\toprule
Commutative Algebra & & Causal-Algebraic Geometry \\
\midrule
Commutative ring $R$ & $\longleftrightarrow$ & Causal algebra $(A, \{{\bf e}_\alpha\}, \leq)$ \\
$\Spec(R)$ & $\sim$ & $\CSpec(A)$ \\
Localization $R_{\mathfrak{p}}$& $\sim$ & Corner algebra ${\bf e}_S A\, {\bf e}_S$ \\
Structure sheaf $\mathcal{O}$ & $\sim$ & Corner algebra presheaf (locality proved; gluing fails) \\
Sheaf cohomology & $\sim$ & Order-complex cohomology \\
$\Spec(\Z)$ & $\sim$ & $\CSpec(k[{\rm Div}(N)])$ (finite truncation) \\
Noetherian condition & $\sim$ & $\gamma \approx 1$ \\
Spectrum determined by ring & $\longleftrightarrow$ & Spectrum is terminal object (Thm.~\ref{thm:uniqueness}) \\
\bottomrule
\end{tabular}
\end{center}
\medskip

\noindent Entries marked $\longleftrightarrow$ are proved
correspondences; entries marked $\sim$ are structural analogies
that motivate the construction but are not claimed as formal
equivalences. In particular, corner algebras play a role analogous
to localizations but are not shown to satisfy the same universal
property.


\subsection{Main theorem}

The results of this paper combine into the following single statement, which we record for ease of citation.

\begin{theorem}[Main Theorem]\label{thm:main}
Let $(C, \leq)$ be a finite poset of width $w$ on $n$ elements.
\begin{enumerate}[nosep]
\item \textup{(Universal property.)} $\CSpec(k[C])$ is the terminal object in the category of causal topologies on $k[C]$: every compatible topology maps to $\CSpec$, the morphism is unique, and no strictly larger compatible topology exists.
\item \textup{(Bridge.)} A subset $S \subseteq C$ is a ring homomorphism open of the structure sheaf if and only if $S$ is causally convex.
\item \textup{(Scaling.)} $\;2^w \leq |\CC(C)| \leq (\lceil n/w \rceil^2 + 1)^w$. Since $|\Int(C)| \geq n^2/(2w)$ for any poset of width $w$ on $n$ elements (Cauchy-Schwarz; proved in Lean~4~\cite{leanrepo}), this gives $\gamma = O(n^{2(w-1)})$ for fixed $w$. A universal lower bound $\gamma \geq 2^w / n^2$ follows from $|\CC| \geq 2^w$ and $|\Int| \leq n^2$.
\item \textup{(Gap.)} If $w = O(1)$, then $\gamma$ is polynomial in $n$. If $w = \Omega(n)$, then $\gamma$ is exponential in $n$.
\end{enumerate}
The Noetherian ratio $\gamma = |\CC(C)|/|\Int(C)|$ is simultaneously a combinatorial invariant (counting convex subsets), a sheaf invariant (counting ring homomorphism opens), and a width detector (polynomial degree determined by width). The irregularity $\kappa = 1 - 1/\gamma$ measures the fraction of non-interval ring homomorphism opens.
All parts are unconditional.
\end{theorem}

\subsection{Open problems}

\begin{enumerate}
\item \textbf{Dimensional scaling.}
The dimension law $\log|\CC([m]^d)| = \Theta(m^{d-1})$ is proved for all $d \geq 2$ in~\cite{difiore2026grid} via a tiling inequality and an ideal/filter injection.
The width scaling law gives $\gamma = O(n^{2(w-1)})$ for fixed~$w$; this bound is sharp on tested examples.
For the $d$-dimensional grid $[m]^d$ on $n = m^d$ elements, the width is $w = m^{d-1} = n^{(d-1)/d}$.
Note that the $O(n^{2(w-1)})$ bound was derived for fixed~$w$; substituting $w \sim n^{(d-1)/d}$ gives a heuristic, not a theorem.
The rigorous result for grids is the dimension law $\log|\CC([m]^d)| = \Theta(m^{d-1})$, proved directly.
Computing exact growth constants $\rho_d$ for $d \geq 3$ and exact formulas for $\gamma([m]^d)$ remain open.
Combined with the upper bound $\log|\CC| \leq w \cdot \log(\lceil n/w \rceil^2 + 1)$ from Theorem~\ref{thm:G}, the dimension law yields a uniqueness result: the entropy exponent $\alpha$ in $\log|\CC| = \Theta(n^\alpha)$ determines the dimension via $d = 1/(1-\alpha)$.
In particular, the Bekenstein--Hawking area law $S \sim m^{d-2}$ for equilibrium configurations implies $d = 4$; see~\cite{difiore2026bh} for details.

\item \textbf{Continuum limit via poset limits.}
Janson~\cite{janson2011} and Hladk\'y, M\'ath\'e, Patel, and Pikhurko~\cite{hladky2015} developed the theory of poset limits (``posons''), in which a sequence of finite posets converges when all finite substructure densities converge.
The natural formulation of the continuum limit for $\CSpec$ is: for a sequence of Poisson processes at increasing density in a Lorentzian manifold, $\gamma$ should be expressible as a functional of the limiting poson, and the $\CSpec$ topology should converge to a pro-$\CSpec$ analogous to Grothendieck's pro-\'etale constructions.
Brightwell and Luczak~\cite{brightwell2015} note that Bombelli's conjecture~\cite{bombelli2000}---that substructure densities distinguish manifold regions up to measure-preserving diffeomorphism---is the poset-limit formulation of the Hauptvermutung.

\item \textbf{Contiguity for $d > 2$.}
For $d = 2$, the uniform measure on 2-dimensional orders is contiguous with a Poisson process in $\mathbb{M}^2$~\cite{elzahar1988, brightwell2008}, providing the only existing proof that a purely combinatorial measure produces manifold-like structure without referencing any continuous geometry.
For $d > 2$, the analogous contiguity is open~\cite{brightwell2015}.
If the gap theorem could be strengthened to show that the \emph{uniform measure on posets with polynomial~$\gamma$} concentrates on manifold-like orders, this would be progress on this open problem.

\item \textbf{Gibbs measures and discrete gravitational dynamics.}
Brightwell and Luczak~\cite{brightwell2012} showed that order-invariant measures on causal sets satisfy equations analogous to the DLR equations of statistical mechanics (Theorem~4.2 of~\cite{brightwell2015}).
If one defines a ``causal Hamiltonian'' $H(C) = -\sum_\alpha \kappa_{\rm BD}(\alpha)$ using the Benincasa--Dowker curvature, the Gibbs measure $e^{-\beta H}$ on the space of finite causal sets is a discrete path integral for gravity.
Extremizing this measure should select posets that extremize total curvature---i.e., solutions to the discrete Einstein equations.
The DLR framework provides the rigorous language; the connection to $\CSpec$ and $\gamma$ would come from expressing the curvature functional in terms of the structure sheaf.

\item \textbf{Causal Grothendieck site.}
Replace the Alexandrov topology with a Grothendieck topology whose covering sieves are families of causal embeddings, potentially yielding non-trivial abelian $H^1$.

\item \textbf{Higher-dimensional scaling.}
Compute $\gamma$ for Poisson sprinklings in $d$-dimensional Minkowski space for $d = 2, 3, 4$ and verify the predicted scaling.

\item \textbf{Join-irreducible generalization.}
The CC Doubling Theorem generalizes: for any finite lattice and any join-irreducible element $j$, adding $j$ doubles $|\CC|$.
This gives a structural decomposition $|\CC(L)| = 2^{j(L)} \cdot C(L)$ where $j(L)$ is the number of join-irreducibles and $C(L)$ accounts for non-join-irreducible corrections, with applications to Boolean lattices, partition lattices, and face lattices of polytopes.
\end{enumerate}


\section*{Acknowledgments}

The formal verification builds on the Mathlib library~\cite{mathlib}.


\begin{thebibliography}{99}

\bibitem{bombelli1987}
L.~Bombelli, J.~Lee, D.~Meyer, and R.D.~Sorkin,
Spacetime as a causal set,
\emph{Phys.\ Rev.\ Lett.} \textbf{59} (1987), no.~5, 521--524.

\bibitem{benincasa2010}
D.M.T.~Benincasa and F.~Dowker,
The scalar curvature of a causal set,
\emph{Phys.\ Rev.\ Lett.} \textbf{104} (2010), 181301.

\bibitem{rota1964}
G.-C.~Rota,
On the foundations of combinatorial theory~I: Theory of M\"obius functions,
\emph{Z.\ Wahrscheinlichkeitstheorie} \textbf{2} (1964), 340--368.

\bibitem{grothendieck1960}
A.~Grothendieck,
\'El\'ements de g\'eom\'etrie alg\'ebrique,
\emph{Publ.\ Math.\ IH\'ES} (1960--1967).

\bibitem{connes1994}
A.~Connes,
\emph{Noncommutative Geometry}, Academic Press, 1994.

\bibitem{voevodsky2003}
V.~Voevodsky,
Motivic cohomology with $\Z/2$-coefficients,
\emph{Publ.\ Math.\ IH\'ES} \textbf{98} (2003), 59--104.

\bibitem{sorkin2003}
R.D.~Sorkin,
Causal sets: Discrete gravity,
in \emph{Lectures on Quantum Gravity} (A.~Gomberoff and D.~Marolf, eds.), Springer, 2003.

\bibitem{malament1977}
D.~Malament,
The class of continuous timelike curves determines the topology of spacetime,
\emph{J.\ Math.\ Phys.} \textbf{18} (1977), 1399--1404.

\bibitem{myrheim1978}
J.~Myrheim,
Statistical geometry,
CERN preprint TH-2538, 1978.

\bibitem{dilworth1950}
R.P.~Dilworth,
A decomposition theorem for partially ordered sets,
\emph{Ann.\ of Math.\ (2)} \textbf{51} (1950), 161--166.

\bibitem{mathlib}
The Mathlib Community,
\emph{Mathlib: The Lean Mathematical Library},
\url{https://github.com/leanprover-community/mathlib4}, 2024.

\bibitem{leanrepo}
T.~DiFiore,
\emph{Causal-Algebraic Geometry in Lean~4},
\url{https://github.com/tomdif/causal-algebraic-geometry-lean}, 2026.

\bibitem{brightwell2015}
G.~Brightwell and M.~Luczak,
The mathematics of causal sets,
arXiv:1510.05612, 2015.

\bibitem{bollobas1991}
B.~Bollob\'as and G.~Brightwell,
Box-spaces and random partial orders,
\emph{Trans.\ Amer.\ Math.\ Soc.} \textbf{324} (1991), 59--72.

\bibitem{brightwell2010}
G.~Brightwell and N.~Georgiou,
Continuum limits for classical sequential growth models,
\emph{Rand.\ Struct.\ Alg.} \textbf{36} (2010), 218--250.

\bibitem{brightwell2008}
G.~Brightwell, J.~Henson, and S.~Surya,
A 2D model of causal set quantum gravity: the emergence of the continuum,
\emph{Class.\ Quantum Grav.} \textbf{25} (2008), 105025.

\bibitem{brightwell1991}
G.~Brightwell and R.~Gregory,
The structure of random discrete spacetime,
\emph{Phys.\ Rev.\ Lett.} \textbf{66} (1991), 260--263.

\bibitem{hawking1976}
S.W.~Hawking, A.R.~King, and P.J.~McCarthy,
A new topology for curved spacetime which incorporates the causal, differential and conformal structures,
\emph{J.\ Math.\ Phys.} \textbf{17} (1976), 174--181.

\bibitem{brightwell2012}
G.~Brightwell and M.~Luczak,
Order-invariant measures on fixed causal sets,
\emph{Comb.\ Prob.\ Computing} \textbf{21} (2012), 330--357.

\bibitem{janson2011}
S.~Janson,
Poset limits and exchangeable random posets,
\emph{Combinatorica} \textbf{31} (2011), 529--563.

\bibitem{hladky2015}
J.~Hladk\'y, A.~M\'ath\'e, V.~Patel, and O.~Pikhurko,
Poset limits can be totally ordered,
\emph{Trans.\ Amer.\ Math.\ Soc.} \textbf{367} (2015), 4319--4337.

\bibitem{bombelli2000}
L.~Bombelli,
Statistical Lorentzian geometry and the closeness of Lorentzian manifolds,
\emph{J.\ Math.\ Phys.} \textbf{41} (2000), 6944--6958.

\bibitem{elzahar1988}
M.~El-Zahar and N.~Sauer,
Asymptotic enumeration of two-dimensional posets,
\emph{Order} \textbf{5} (1988), 239--244.

\bibitem{rideout2000}
D.~Rideout and R.D.~Sorkin,
A classical sequential growth dynamics for causal sets,
\emph{Phys.\ Rev.\ D} \textbf{61} (2000), 024002.

\bibitem{barnette2019}
B.~Barnette, W.~Nichols, and T.~Richmond,
The number of convex sets in a product of totally ordered sets,
\emph{Rocky Mountain J.\ Math.} \textbf{49} (2019), no.~2, 369--385.

\bibitem{difiore2026grid}
T.~DiFiore,
Order-convex subsets of grid posets: A new exponential combinatorial class,
preprint, 2026.

\bibitem{difiore2026bh}
T.~DiFiore,
Black hole thermodynamics from counting convex subsets of a lattice,
preprint, 2026.

\bibitem{edelmanjamison1985}
P.~H.~Edelman and R.~E.~Jamison,
The theory of convex geometries,
\emph{Geom.\ Dedicata} \textbf{19} (1985), 247--270.

\bibitem{lihova1999}
J.~Lihov\'a,
On convex sublattices of lattices,
\emph{Algebra Universalis} \textbf{42} (1999), 265--294.

\bibitem{spiegel1997}
E.~Spiegel and C.~J.~O'Donnell,
\emph{Incidence Algebras},
Marcel Dekker, New York, 1997.

\end{thebibliography}

\end{document}
